---
title: "APIs & Web Scraping for Medical & Public Health Research"
format:
  html:
    theme: cosmo
    toc: true
    number-sections: false
execute:
  echo: true
  warning: false
  message: false
---
# Packages
```{r}
# ---- setup: install + load everything we use ----
if (!requireNamespace("pacman", quietly = TRUE))
  install.packages("pacman")

pacman::p_load(
  # tidy data / grammar of data
  tidyverse,        # ggplot2, dplyr, tidyr, readr, purrr, tibble, stringr, forcats
  dplyr, tibble,    # (listed explicitly in some chunks)
  purrr, stringr,
  janitor,           # clean_names() and other data cleaning helpers
  
  # HTTP / API helpers
  httr2, httr, jsonlite,
  
  # Census
  tidycensus, censusapi,
  
  # NHANES
  NHANES, nhanesA,
  
  # Tables & formatting
  flextable, scales,
  
  # Misc (reading local CSVs in join example)
  readr,

  # Web scraping
  rvest
)
```

# Welcome + goals

**Today you’ll learn to:**

- Explain what an API is and why researchers use them
- Make API requests and parse JSON into analysis-ready tables using R and Python
- Use public health / biomedical APIs (including the Census API)
- Understand when web scraping is appropriate and how to do it responsibly
- Scrape a simple site; understand authenticated scraping patterns (safely)

**Why it matters**

- Reproducible data retrieval 
- Faster updates and larger-scale studies
- Cleaner provenance (query + timestamp + version)

## Agenda

1. API fundamentals
2. API workflow demonstrations  
3. Public health & biomedical APIs  
4. Census API example  
5. Web scraping concepts + ethics  
6. Reproducibility checklist + resources

## Ground rules (research integrity + compliance)

- Prefer **official APIs** and **bulk downloads** over scraping
- Respect **terms of use**, **rate limits**, and institutional policy
- Avoid collecting **PII/PHI**
- Use **caching** and **delays** to minimize server load
- Document provenance: *what you requested, when, from where*

# Part 1 — APIs

## What is an API?

An **Application Programming Interface (API)** is a structured way to request data from a service.

For researchers, APIs are “programmable search + download”:

- predictable formats (often **JSON** but sometimes **XML** or **CSV**) 
- documented inputs (parameters)
- standardized outputs (fields)

## Common data formats: CSV, JSON, and XML

APIs commonly return data in **JSON** or **XML**, while researchers often analyze data in tabular formats like **CSV**.

### CSV (Comma-Separated Values)

A CSV is a plain-text table:
- rows = observations
- columns = variables
- easy to open in Excel, R, Python, or Stata

Example:

```csv
id,age,sex
101,42,F
102,35,M
```

### JSON (JavaScript Object Notation)

JSON stores data as nested key–value structures:

- flexible and web-friendly
- commonly used in modern APIs
- may require “flattening” into tables

Example:

```json
[
  {
    "id": 101,
    "age": 42,
    "sex": "F"
  },
  {
    "id": 102,
    "age": 35,
    "sex": "M"
  }
]
```

### XML (eXtensible Markup Language)

XML organizes data using nested tags:

- older but still widely used in biomedical and government systems
- highly structured and descriptive
- commonly seen in legacy APIs and document-based databases

Example:

```xml
<participants>
  <participant>
    <id>101</id>
    <age>42</age>
    <sex>F</sex>
  </participant>
  <participant>
    <id>102</id>
    <age>35</age>
    <sex>M</sex>
  </participant>
</participants>
```

### Research workflow

A common workflow is:

API request → JSON/XML response → parse/clean → tabular dataset (CSV/data frame)

## Why APIs are preferred

- **Reproducible**: same query, same retrieval method
- **More stable**: APIs change less often than website layouts
- **Automatable**: supports batching, scheduling, and pipelines
- **Transparent**: explicit parameters, documentation, and usage rules

### What is pagination?

**Pagination** is how an API delivers large result sets in smaller “pages” to control load and response size.

Why pagination exists:

- prevents oversized responses
- improves reliability
- supports rate limiting


## Core vocabulary (fast)

- **Endpoint**: the URL for a resource (e.g., `/search`)
- **Method**: `GET` (retrieve), `POST` (submit query)
- **Parameters**: filters like `?year=2023&state=CT`
- **Auth**: key/token, sometimes OAuth
- **Rate limits**: requests per time period
- **Pagination**: page/offset/cursor for large results
- **Response**: JSON objects you parse into tables

## RESTful APIs in biomedical/public health research

Many health and public data services expose **RESTful** (or “REST-like”) APIs. REST isn’t a single technology; it’s a design style that makes APIs predictable and interoperable.

### The core idea: resources + representations

In REST-style APIs, you interact with **resources** (things) using URLs, and you receive **representations** (often JSON) of those resources.

Examples of “resources” in biomedical/public health contexts:
- a publication record
- a clinical trial
- an adverse event report
- a county-level indicator
- a dataset row in an open-data portal

### Common HTTP methods (what they *mean*)
- `GET` — retrieve data (most common for research data pulls)
- `POST` — submit a query or create a record (common for complex searches; also for systems like REDCap when writing data)
- `PUT/PATCH` — update a record (more common in operational systems than public datasets)
- `DELETE` — remove a record (rare in public data APIs)

**Research norm:** you’ll primarily use `GET`, and occasionally `POST` for “search” endpoints or authenticated systems.

### Status codes you will see (and how to respond)
- `200 OK` — success
- `400 Bad Request` — your parameters are malformed
- `401 Unauthorized` — missing/invalid token
- `403 Forbidden` — you’re authenticated but not allowed (permissions/terms)
- `404 Not Found` — wrong endpoint or resource ID
- `429 Too Many Requests` — you hit a rate limit (slow down; retry later)
- `500/503` — server problems (retry with backoff)

### Headers (metadata about the request/response)
Headers often matter in practice:
- `Accept: application/json` — ask for JSON
- `User-Agent` — identify your tool (polite practice for scraping; sometimes for APIs too)
- `Authorization: Bearer <token>` — common token format
- API key headers (varies by service)

### Parameters and filtering
Most APIs let you filter/shape results using query parameters, such as:
- date ranges: `start_date=...&end_date=...`
- geography: `state=09&county=001`
- selecting fields: `fields=...`
- sorting: `sort=...`

### Data formats: JSON and “tidying”
API responses are often **nested JSON**, meaning you may need to:
- “flatten” nested objects into columns
- unpack arrays (one-to-many relationships)
- cast strings to dates/numbers
- handle missing fields robustly

### Provenance: what to record for methods/reproducibility
For every API pull, log:
- endpoint URL
- parameters (including filters and selected fields)
- access date/time (UTC)
- API version (if available)
- software/package versions used to retrieve the data

## How biomedical researchers use APIs (examples)

- Literature: retrieve PubMed metadata for reproducible searches
- Safety: monitor OpenFDA adverse event reports
- Trials: query ClinicalTrials.gov trial registries
- Surveillance: download public health indicators and dashboards
- Context: integrate Census covariates into analyses

## API workflow (the “researcher loop”)

1. Identify source + documentation  
2. Draft a query (fields, filters, time range)  
3. Request data (respect limits)  
4. Validate + clean  
5. Save raw + processed data  
6. Record provenance (URL, params, date, version)  

## Demonstration approach today (language-agnostic)

We’ll show examples in **R and Python** patterns:
- build URL with params
- GET request
- parse JSON
- transform to a table
- export to CSV
- log provenance

# Part 2 — Publicly available APIs (tour)

## Public health / biomedical APIs you can use

Examples to explore (depending on your research):

- PubMed / NCBI E-utilities (literature metadata)
  - https://www.ncbi.nlm.nih.gov/books/NBK25500/
- ClinicalTrials.gov (trial registry metadata)
  - https://clinicaltrials.gov/data-api/about-api
- OpenFDA (drugs, devices, adverse events)
  - https://open.fda.gov/apis/
- CDC open data (often via Socrata patterns)
  - https://wonder.cdc.gov/wonder/help/wonder-api.html
- Census API (ACS, decennial, geography)
  - https://www.census.gov/data/developers/data-sets.html 


## APIs in the tools you already use (REDCap | R packages | Python libraries)

### REDCap APIs (download *and* upload)

Many institutions run **REDCap**, which includes an API for:

- exporting data (CSV/JSON)
- importing/updating records
- exporting metadata (data dictionaries)
- managing instruments/events (project-dependent)

In practice, this enables reproducible pulls of study data *and* automated pipelines for cleaning/QA workflows.

### R packages that provide API-style access

In R, many packages act as friendly clients to public data APIs or maintained data endpoints. Examples you may see in health research workflows include:

- `cdcfluview` (influenza surveillance-related retrieval patterns)
- `RNSSP` (syndromic surveillance access patterns, where available/authorized)
- `NHANES` (NHANES data retrieval and convenience functions)

When using package-based access, still apply the same reproducibility habits:

- record package version
- record query parameters
- record access date/time and dataset version


## Choosing an API dataset (quick criteria)

- What is the **unit of analysis**? (person, county, hospital, trial)
- How often is it updated?
- What are the **terms/limits**?
- What fields are available? Are definitions clear?
- Are there stable identifiers? (FIPS, NPI, trial IDs, DOIs)


## API demonstrations (one per source)

### NHANES R package — survey data access

**Use cases**
- quick access to NHANES data
- access the most recent survey cycles without manual file downloads
- merge datasets across modules using participant ID (`SEQN`)

**NHANES organizes data into separate modules (demographics, labs, examinations, questionnaires, etc.), which are linked by participant ID (`SEQN`).**

#### Query demographic and body measurement datasets
```{r}
# R demo skeleton (NHANES via nhanesA package)
# packages used: NHANES, nhanesA, tidyverse

# Example: get demographic data for 2021-2023 cycle (most recent) denoted by "L"
nhanes_demo <- nhanes('DEMO_L')
#head(nhanes_demo)

# Example: get demographic data for 2021-2023 cycle (most recent) denoted by "L"
nhanes_bmx <- nhanes('BMX_L')
#head(nhanes_bmx)
```

The syntax changes across APIs, but the overall research workflow is remarkably similar.

#### Then you can merge/join datasets using the participant ID (`SEQN`) to create an analysis-ready table.

It is important to note that using a data dictionary is crucial to understand the variables and their coding. For example, `RIAGENDR` is coded as 1 for male and 2 for female.

```{r}
nhanes_data <- nhanes_demo %>%
  left_join(nhanes_bmx, by = "SEQN") %>%
  select(SEQN, RIDAGEYR, RIAGENDR, BMXBMI)

head(nhanes_data)
```

### PubMed (NCBI E-utilities) — literature metadata

**Use cases**

- reproducible literature searches
- retrieve PMIDs + basic citation metadata
- build screening datasets

**NCBI requests should follow usage guidelines and include reasonable request pacing for large retrieval jobs.**

```{r }
# R demo skeleton (E-utilities via httr2)

# packages used: httr2, jsonlite, dplyr, tibble

base <- "https://eutils.ncbi.nlm.nih.gov/entrez/eutils/esearch.fcgi"
query <- list(
  db = "pubmed",
  term = "asthma[MeSH Terms] AND 2023[pdat]",
  retmode = "json",
  retmax = 20
)

resp <- request(base) |> req_url_query(!!!query) |> req_perform()
txt <- resp |> resp_body_string()
dat <- fromJSON(txt)

pmids <- dat$esearchresult$idlist
tibble(pmid = pmids)
```

### ClinicalTrials.gov — trial registry metadata

**Use cases**

- find trials by condition/intervention/location
- compare outcomes reporting patterns
- sponsor/phase landscape scans

```{r}
# R demo skeleton (ClinicalTrials.gov via httr)
#  
# Packages used: httr, jsonlite, dplyr, tibble

# ---- endpoint & parameters ----
url <- "https://clinicaltrials.gov/api/v2/studies"

params <- list(
  `query.term` = "diabetes",
  pageSize      = 10
)

# ---- request ----
resp <- GET(url, query = params, timeout(30))

# error if HTTP status is not 200
stop_for_status(resp)

# ---- parse JSON ----
data_txt <- content(resp, as = "text", encoding = "UTF-8")
data     <- fromJSON(data_txt, flatten = TRUE)

# ---- extract the studies array ----
studies <- data$studies            # flatten nested JSON into a data frame

# ---- inspect / use ----
studies_tbl <- as_tibble(studies)
#print(studies_tbl, n = 10, width = Inf)
#str(studies_tbl)
``` 

### OpenFDA — adverse events / labeling

**Use cases**

- pharmacovigilance exploration
- counts over time, stratified by drug/product fields
- teaching filtering + pagination

```{r}
# R demo skeleton (OpenFDA via httr)

# packages used: httr, jsonlite, dplyr, tibble

# ---- endpoint & parameters ----
base   <- "https://api.fda.gov/drug/event.json"
params <- list(
  search = 'patient.drug.medicinalproduct:"metformin"',
  limit  = 5
)

# ---- request ----
resp <- GET(base, query = params, timeout(30))
stop_for_status(resp)                       # error if HTTP status ≥ 400

# ---- parse JSON ----
txt <- content(resp, as = "text", encoding = "UTF-8")
j   <- fromJSON(txt, flatten = TRUE)

# ---- extract the results array ----
fda_df <- j$results %>%            # NULL if no hits
        as_tibble()

#print(fda_df, n = 5, width = Inf)  # show the 5 rows we requested
```

### CDC open data (Socrata-style)

Many government open-data portals use Socrata APIs, which support SQL-like query parameters such as `$select`, `$where`, and `$limit`.

https://dev.socrata.com/docs/endpoints 

**Use cases**

- quick access to many public datasets
- teach “select/where/limit” query patterns

**Run this chunk to find the endpoint for a dataset of interest (e.g., COVID-19 hospitalization surveillance)**

Search other datasets by changing the `q` parameter and/or domain filter. Note that some CDC datasets are on data.cdc.gov, but others may be on different domains (e.g., wonder.cdc.gov). You can also explore the Socrata API documentation for more complex queries and dataset discovery.

```{r}
# R demo skeleton: Socrata dataset discovery (you will replace the query with your own topic of interest)

# packages used: httr2, jsonlite, dplyr, tibble

catalog <- request("https://api.us.socrata.com/api/catalog/v1") |>
  req_url_query(q = "COVID-19 hospitalization", domains = "data.cdc.gov") |>
  req_perform() |>
  resp_body_json()

catalog$results[[1]]$resource$id  
```

**Run this chunk with the dataset-specific endpoint to get data (replace the XXXX-XXXX with the one you found in the previous step)**

```{r}
# R demo skeleton: Socrata-style endpoint (you will replace with a specific dataset)

# packages used: httr2, jsonlite, dplyr, tibble

# Many Socrata datasets support SODA queries via $select, $where, $limit.
# Replace with a concrete dataset endpoint you choose for class:
# Example form: https://data.cdc.gov/resource/XXXX-XXXX.json

endpoint <- "https://data.cdc.gov/resource/akn2-qxic.json"

resp <- request(endpoint) |>
  req_url_query(`$limit` = 10) |>
  req_perform()

resp_dat <- fromJSON(resp_body_string(resp), flatten = TRUE)
as_tibble(resp_dat)
```

# Part 3 — Census API (token + joins)

## Why the Census/ACS matters in health research

Adds **population context** for: 

- Denominators to turn counts into rates
- Social determinants proxies (income, education, housing, internet)
- Fine-grained geographies (tract, ZIP, county, ZCTA, PUMA, etc.)
- Annual or 5-year series for trend analysis

Common outputs: 

- tract/county/state estimates 
- ACS variables (multi-year)

## Getting a Census API token

1. Request one free key: https://api.census.gov/data/key_signup.html
2. Store it safely (e.g. in ~/.Renviron as CENSUS_API_KEY=ABC…).
3. Load it in R:

## Store your Census key (R example)
```{r}
# Set your key in .Renviron (permanenet, editsable, one-time setup)
# Should look like this: CENSUS_API_KEY=YOUR_KEY_HERE

#file.edit("~/.Renviron")
```

## Load your Census API key from environment variables (R example)
```{r}
# You only need to run this once per R session to load the key into your environment. It will be available for all tidycensus functions after this.

# packages used: tidycensus

#Sys.getenv("CENSUS_API_KEY")
census_api_key(Sys.getenv("CENSUS_API_KEY"), install = TRUE, overwrite = TRUE)
```

## Test your key and make a simple request (R example)
```{r}
# This is a simple test to see if your key is working and to get a basic population estimate for all states in 2023. You should see a table with state names, population estimates, and margins of error.

# packages used: tidycensus

get_acs(geography = "state", variables = "B01001_001", year = 2023)
```

## Census API: geographies + joins 

### Guide to US Census Bureau Data Tables (ACS)

The American Community Survey (ACS) tracks over 1,400 demographic and economic metrics. Instead of searching blindly, you can decipher any Census table code instantly using its alphanumeric structure.

#### 1. Anatomy of a Census Variable Code

Take the code **`B01001_001`**:
* **`B`** (or **`C`**): The Table Type (*Base* or *Collapsed*).
* **`01`**: The Broad Subject Code (*Age & Sex*).
* **`001`** (before the underscore): The unique table sequence sequence number.
* **`_001`** (after the underscore): The specific **line number** (row) inside that table. 
  * *Tip: Line `_001` is almost always the "Total" count for that whole table.*

#### 2. Cheat Sheet: Table Subject Prefixes

The first two digits after the leading letter tell you exactly what topic the table covers:

| Prefix | Subject Category | Example Metrics Included |
| :--- | :--- | :--- |
| **`01` to `07`** | **Core Demographics** | Age, Sex, Race, Ancestry, and Place of Birth |
| **`08`** | **Commuting & Transit** | Remote work, public transit use, and travel times to work |
| **`11` to `13`** | **Households & Family** | Marital status, family size, and fertility rates |
| **`15` to `16`** | **Education & Language** | High school/college graduation rates, languages spoken at home |
| **`17` to `23`** | **Economics & Finance** | Median Household Income, poverty status, and employment data |
| **`24`** | **Occupations** | Employment fields (Tech, Manufacturing, Service industry splits) |
| **`25`** | **Housing & Living** | Rent vs. Own, home values, monthly utility costs, vehicle counts |

**Helpful website to search tables** https://censusreporter.org

#### 3. High-Engagement Tables for Student Projects

When querying datasets using `get_acs()`, these tables are the most popular for economic, sociological, and geographical data analysis:

* **`B19013`** — Median Household Income
* **`B17001`** — Poverty Status by Sex and Age
* **`B15003`** — Educational Attainment (Population 25 years and older)
* **`B25077`** — Median Value of Owner-Occupied Housing Units
* **`B28002`** — Presence and Type of Household Internet Access (The "Digital Divide")
* **`B25044`** — Vehicles Available per Household

#### 4. `B` Tables vs. `C` Tables (Base vs. Collapsed)
* **`B` (Base Tables):** Provide micro-detailed row breakdowns (e.g., tracking poverty across 20+ micro-age brackets).
* **`C` (Collapsed Tables):** Take the exact same data but compress it into fewer, broader rows (e.g., grouping that same poverty data into just 3 wide age buckets). 
  * *If a `B` table dataset feels overwhelming, change the first letter to a `C` for a cleaner overview!*

```{r}
# This example shows how to get population estimates by county in Connecticut, calculate the percentage of the state population for each county, and then build a nice table to display the results. You can adapt this pattern to other geographies, variables, and states as needed.

# packages used: tidycensus, dplyr, flextable, scales

# Load your Census API key (if you haven’t already)
census_api_key(Sys.getenv("CENSUS_API_KEY"))

# ── 1. Population by County (Connecticut) ────────────────────────────────────

county_pop <- get_acs(
  geography = "county",
  variables = "B01001_001",  # we use Variables to calucalte the percentage of the state population for each county; this is the total population variable in the ACS
  state = "CT",
  year = 2023,
  survey = "acs5"
) %>%
  mutate(
    NAME = gsub(" County, Connecticut", "", NAME),
    pct_of_state = (estimate / sum(estimate)) * 100
  )

county_pop
```

## Now we can build a nice table to display the population estimates by county, along with the margin of error and the percentage of the state population.

```{r}
build_county_pop_table <- function() {
  county_pop %>%
    arrange(desc(estimate)) %>%
    select(
      County = NAME,
      Population = estimate,
      #MOE = moe,
      `% of State` = pct_of_state
    ) %>%
    mutate(
      Population = comma(Population),
      #MOE = comma(MOE),
      `% of State` = paste0(round(`% of State`, 1), "%")
    ) %>%
    flextable() %>%
    set_header_labels(
      County = "County",
      Population = "Population",
      #MOE = "Margin of Error",
      `% of State` = "% of State"
    ) %>%
    theme_vanilla() %>%
    bold(part = "header") %>%
    bg(part = "header", bg = "white") %>%
    color(part = "header", color = "black") %>%
    bg(i = seq(2, nrow(county_pop), 2), bg = "#F2F2F2") %>%
    align(j = c("Population", "% of State"), align = "right", part = "all") %>% #removed "MOE" since its all blanks
    set_table_properties(width = 1, layout = "autofit") %>%
    add_footer_lines("Source: U.S. Census Bureau, ACS 5-Year Estimates 2023. Table B01001.")
}

build_county_pop_table()
```

## Now you can call the `build_county_pop_table()` function to display the formatted table of population estimates by county in Connecticut.

```{r}
# Call this to display the table:
build_county_pop_table()
```


# Wrapping it all up – From curiosity to pipeline

## 1. Common skeleton

- Identify endpoint + docs
- Auth if needed
- Build query (params/body)
- Send request (GET/POST)
- Handle errors / rate limits
- Parse JSON/XML → table
- Clean, validate, record provenance

## 2. Reproducibility checklist

☐ Save raw response (or hash) 

☐ Log URL/params + timestamp + API version 

☐ Pin package versions / environment info 

☐ Respect robots.txt / rate limits 

☐ Secure tokens (env vars or vault) 

## 3. Where to go next

- Explore additional endpoints: FDA Device, CMS Open Payments, EPA, WHO, UN, etc.
- Learn asynchronous / parallel request patterns for large pulls (furrr, async in Python).
- Pair APIs with geospatial analysis (sf, tigris, leaflet).
- Automate updates with cron / GitHub Actions.

## 4. Key takeaway

- Different APIs, one mental model: endpoints + params + auth + pagination + parse + provenance. Master the pattern once and you can unlock a universe of public-interest data for transparent, up-to-date, and scalable health research.

- Feel free to drop these enhanced sections into your materials and tweak the wording or examples to match your audience and time constraints.

# Web scraping

## What is web scraping?

Web scraping is the process of programmatically extracting data from websites. It’s often used when:
- No API is available
- The API is too limited or slow for your needs
- You want to capture data that’s only presented in HTML format

## Today we will be scrapping a Wikipedia page to extract gloabal COVID cases and deaths by country

### STEP 1 – Open the page & inspect it

- Visit the **https://en.wikipedia.org/wiki/COVID-19_pandemic_by_country_and_territory** (Chrome/Edge/Firefox).
- Right-click anywhere inside the data table → Inspect (or “Inspect Element”).
- Scroll through DevTools and see where the table is highlighted on the webpage

### STEP 2 – Identify the table in the HTML structure

- The table is nested inside a `<div>` with class `mw-parser-output`, which contains the main content of the Wikipedia page.
- The table itself is a `<table>` element with class `wikitable sortable`.
- when you find the right element, right click on the `<table>` tag in DevTools → Copy → Copy Full XPath

**Here is he xpath example for the table** 

/html/body/div[3]/div/div[3]/main/div[3]/div[3]/div[2]/div[24]/table

### STEP 3 – Use R to read the HTML and extract the table

```{r}
library(rvest); library(janitor); library(dplyr)

url <- "https://en.wikipedia.org/wiki/COVID-19_pandemic_by_country_and_territory" # Define the target Wikipedia URL as a text string
page <- read_html(url) # Download the entire HTML structure of the web page into R memory

covid_tbl <- page %>%
  html_element(xpath = "/html/body/div[3]/div/div[3]/main/div[3]/div[3]/div[2]/div[24]/table") %>% # Use the XPath to locate the specific table element within the HTML structure
  html_table(fill = TRUE) %>% # Convert the HTML table into an R data frame, filling in any missing cells with NA
  as_tibble(.name_repair = "unique") %>% # Convert the data frame to a tibble for easier manipulation, ensuring that column names are made unique if there are duplicates
  clean_names() %>% 
  select(-x1)

head(covid_tbl)
```

## This works, but it’s not very reusable. If we want to scrape another table on the same page, we’d have to copy-paste and change the XPath again. Let’s make a function to do this more cleanly.

```{r}
library(rvest); library(janitor); library(dplyr)

# --- New Code Chunk: Function to Scrape Tables via XPaths ---

url <- "https://en.wikipedia.org/wiki/COVID-19_pandemic_by_country_and_territory" # Define the target Wikipedia URL as a text string
page <- read_html(url) # Download the entire HTML structure of the web page into R memory

# 1. Define the custom function
scrape_by_xpath <- function(html_page, xpath_string) { 
  html_page %>%
    html_element(xpath = xpath_string) %>%
    html_table(fill = TRUE) %>%
    as_tibble(.name_repair = "unique") %>%
    clean_names()
}

# 2. Store your two specific paths in variables
path_cases    <- "/html/body/div[3]/div/div[3]/main/div[3]/div[3]/div[2]/div[24]/table"
path_timeline <- "/html/body/div[3]/div/div[3]/main/div[3]/div[3]/div[2]/div[30]/table"

# 3. Use the function to scrape both tables seamlessly
covid_tbl_1 <- scrape_by_xpath(page, path_cases) %>% select(-x1) # drops the empty index column
covid_tbl_2 <- scrape_by_xpath(page, path_timeline) %>% select(-x1) # drops the empty index column

# 4. Preview the tables
head(covid_tbl_1)
head(covid_tbl_2)
```

# Wrapping Up 

## 1. Key take-aways

- **Same playbook, different sites**  
  No matter where the data lives, the steps are: find the URL, send a
  request, get the reply, turn it into a table, and write down what you did.

- **Keep it reproducible**  
  Save the URL (plus any filters), note the date you downloaded, and jot down
  which R packages you used.  That’s enough for someone else (or future-you)
  to rerun the pull.

- **Respect the ecosystem**  
   Use an official API when one exists, slow down big jobs, and read the “terms
  of use.”  Scraping is the last resort.

- **Automate early**  
   Wrap your code in a function and schedule it—then fresh data is only a click
  (or a cron job) away.

- **Stay curious**  
  New health data sources pop up all the time.  Once you know the basics,
  trying a new one is mostly copy-paste.

## 2. Suggested next steps

| Level-up | Action |
|----------|-----------------|
| **Async / parallel pulls** | Try `furrr` or `curl_multi()` for paginated APIs with thousands of calls. |
| **Secrets management** | Store tokens in `.Renviron`, `keyring`, or a cloud vault; never hard-code in scripts. |
| **Robust scraping** | Explore `chromote`, `RSelenium`, or `playwright` for JavaScript-heavy sites. |
| **Geospatial joins** | Pair API outputs with `sf`, `tigris`, or `tidycensus::get_decennial()` boundary data. |
| **Publication-ready docs** | Convert this notebook to PDF, slide decks, or dashboards with Quarto multi-format. |

## 3. Handy links

| Link     |   Why you’d use it    |
|----------|-----------------------|
| **CT Open Data Portal** — <https://data.ct.gov> | State-level health & policy datasets in a Socrata wrapper (supports `$select`, `$where`, etc.). |
| **Census Developer Portal** — <https://www.census.gov/data/developers/data-sets.html> | Official docs, variable lists, and examples for ACS, Decennial, Economic Census, and more. |
| **NCHS Vital Statistics** — <https://www.cdc.gov/nchs/data_access/vitalstatsonline.htm> | Birth/death micro-data, natality, mortality, and compressed mortality files (often FTP + API hybrids). |
| **ClinicalTrials.gov API** — <https://clinicaltrials.gov/api/info/study_fields_list> | Retrieve study fields, statuses, locations; essential for trial landscape or compliance work. |
| **OpenFDA** — <https://open.fda.gov/apis/> | Drug, device, food event surveillance; supports powerful nested queries and pagination. |
| **NCBI E-utilities** — <https://www.ncbi.nlm.nih.gov/books/NBK25500/> | Programmatic PubMed, PMC, GenBank access; cornerstone for reproducible literature searches. |
| **Socrata API docs** — <https://dev.socrata.com/docs/endpoints> | Generic guide for thousands of city/state portals (CDC, HHS, EPA, etc.). |
| **Census main API landing** — <https://api.census.gov/data/> | Direct links to every vintage & survey; great for discovering obscure datasets (e.g., LEHD, CPS). |
| **CDC WONDER API help** — <https://wonder.cdc.gov/wonder/help/wonder-api.html> | Documentation for WONDER’s SQL-like query interface (mortality, natality, underlying cause of death). |
| **OHDSI GitHub** — <https://github.com/OHDSI/> | Open-source tooling (Atlas, Achilles) around the OMOP common data model — valuable for EHR research. |
| **OHDSI Athena vocabularies** — <https://athena.ohdsi.org/search-terms/start> | Download standard vocabularies (SNOMED, RxNorm, LOINC) for mapping local codes to OMOP. |
| **Example CDC dataset** — <https://dev.socrata.com/foundry/data.cdc.gov/wnjn-mave> | A live schema page showing field types, sample API calls, and metadata. |

## 4. Thank you & stay connected!

Please email me at maximilian.wegener@yale.edu if you have any questions.